Relationships in the extremes and their influence on competition and coexistence

### 2022-03-20 paper/Paper.Rnw
responded to Dan's commented and moved commented out intro to Scraps.txt
adding to discussion a bit

%***DAN: Go through and make sure all citation callouts list papers in date order, oldest to newest,
%except when multiple pubs by the same first author are listed they are all together.

@@

%make the above abstract ideas more concrete - 
%***DAN: This para commented out for length, added a sentence about the systems to the above paragraph to take the place of this text
%To make the above ideas more concrete, we briefly mention examples from prior work to 
%which the ideas may apply. 
%\cite{Chesson_2012} provides two examples of competitive plant communities in which different species
%are likely influenced differently by phenologically distict environmental variables: an Australian 
%grass-forb community, and an Arizona community of desert annuals. 
%The communities follow the criteria required for temporal storage effects to operate \citep{Chesson_2012}. In both 
%examples, distinct species have different periods of germination, competition is influenced by resource 
%consumption, and population growth is buffered by overlapping generations or via seed banks.
%Temporal differences between species in germination rates suggest the possibility that 
%population growth of distinct species may be driven by rainfall signals in distinct parts
%of the growing season. Thus all the ingredients seem to be present by which climatic changes may
%alter ATAs between environmental variables which are relevant for the coexistence
%of the example species, motivating our efforts, here, to perform a detailed study of how ATAs can influence competition.

%Review of lit on changes in means and variances of climate signals and associated impacts on coexistence
%**DAN: this para commetned out for length, added a couple of sentences to an aboce para to substitute
%Although, to our knowledge, the potential importance of climatic changes in ATAs for species coexistence has 
%not been studied, the importance for coexistence of changes 
%in means and variability of environmental variables has been studied in many systems. We briefly 
%review some of that literature.
%\cite{diPaola_2012} concluded that coexistence of drought sensitive,
%flood resistant species with flood sensitive, drought resistant species in 
%Mediteranean coastal swamp forests of central Italy would be jeopardized
%by projected changes in mean rainfall. 
%\cite{Descombes_2020} found that if warming trends promote faster expansion toward 
%higher altitudes of herbivore populations than of plant populations, newly arrived
%herbivores at high altitudes can inflate plant species richness by reducing dominant plant species 
%in favor of subordinates. Contrastingly, \cite{Usinowicz_2021} concluded, also for
%alpine plants, that upward range expansion under warming trends would be good for the growth and survival of 
%all species they considered, but that changes in competition would more than offset these direct benefits. 
%Numerous studies take an ``event-focused'' 
%perspective \citep{Jentsch_2007}, examining how increased climatic variability, in the
%form of extreme weather events, can alter competitive interactions. For instance, 
%\cite{White_2001} showed that single extreme heat events can alter competition 
%between C3 and C4 plants. And \cite{Parepa_2013} showed that extreme variability in soil
%nutrient availability for plants, mediated by soil moisture content, can alter 
%the potential for invasive species to enter an ecosystem and establish, a process that involves
%the invader overcoming competition from resident species. 
%Using MCT, it has been demonstrated that increased variability can promote coexistence in plant 
%communities, via storage effects \citep{Adler_2006,  Angert_2009}.

%***DAN: Jasmin, I commented the below sentence because that body of work is famously 
%complex and sophisticated, and I did not feel we could do it justice in short space.
%decreased growth season precipitation can decrease the competitive ability of dominant 
%forest tree species \citep{Clark_2011}, 
%***DAN: Jasmin, I commented the below sentence because I'm not sure the two studies 
%cited there are actually directly about coexistence, in the MCT sense. One of them is
%a general review on invasive species, and the other has to do with competition,
%but coexistence is not mentioned. I searched both papers for "coexist" and found
%no hits. OK, both are relevant, but maybe not so directly, to what we are talking about,
%and since this paragraph could anyway stand to be shortened a bit, I took these out.
%Alternatively, dominant species exclusion may be beneficial as rare species can increase 
%in diversity \citep{Mariotte_2013}, while an increase in dominant species fitness can 
%reduce community diversity, typical of invasive species \citep{Vitousek_1997}.

@@

%***DAN: I think we should replace the terminology "invasion growth rate" and "IGR" by "growth when rare" and "GWR",
% throughout the paper. Has to be done in the main text and sup mat and the figures themselves.
% Need also search for "invader" since sometimes we use that word. Also think of any other related terminology that needs to 
% be altered as part of the overall change.

@@

%G4
To achieve goal G4 from the Introduction, i.e., to provide an empirical example of species 
coexistence which highlights the role ATAs can play, we use a laboratory chemostat
system of freshwater diatoms explored by \cite{Descamps_2005}.
Those authors used measurements of growth rates of \emph{Fragilaria crotonensis} and 
\emph{Cyclotella pseudostelligera} over a range of temperatures to parametrize a resource
competition model. The model is 
\begin{align}
\frac{dS}{dt} &= D(S_0 - S) - Q_1x_1 \frac{V_1 S}{K_1 + S} - Q_2 x_2 \frac{V_2 S}{K_2 + S} \\
\frac{dx_j}{dt} &= x_j \frac{V_j S}{K_j+S}-D x_j, 
\end{align}
for $j=1,2$, where $S$ is a extracellular silicate concentration in the chemostat, $x_1$ is the population density
of \emph{F. crotonensis} and $x_2$ is that of \emph{C. pseudostelligera}, 
$D$ is the chemostat outflow rate, $S_0$ is concentration of silicates in the chemostat
inflow, the $Q_j$ are amounts of silicate per cell, the $V_j$ are maximum population growth rates, and
the $K_j$ are half-saturation constants for the dependence of growth rates on nutrient concentration.
The temperature-dependent parameters $Q_j$, $V_j$, and $K_j$ were obtained by 
\cite{Descamps_2005} through batch experiments and curve fitting or interpolation.
\cite{Descamps_2005} showed that permanent coexistence of the two species occurs when 
temperature fluctuates periodically, 
$\theta(t)=\theta_0 +  a\sin(2\pi t/P)$, 
with parameters $\theta_0 = 18^\circ$C, $a = 6^\circ$C, $P = 60$d. 
\cite{Ellner_2016, Ellner_2019} broke down coexistence into contributing mechanisms
according to (\ref{eq:previous_decomp}) for these same values of
$\theta_0$, $a$ and $P$. We further decomponse
coexistence according to (\ref{eq:full_decomp}), and we do so for ranges of values
of $\theta_0$, $a$ and $P$. In contrast to the lottery model case, for which simplifying assumptions
made it sufficient to decompose only the GWR of species 1, for this
%***DAN: Please search for "growth when rare" and replace with GWR (except for the first instance,
%where GWR is defined). If one defines an acronym, one then has to use it.
example we considered $\olr_{i \bs i} -q_{ij}\olr_{j \bs i}$ for both $i=1$, $j=2$ and for
$i=2$, $j=1$. Further details of the model setup are in SI section \ref{SI-sect:diatom_model}
and specifics of how our decomposition extends to this empirical case are in 
SI section \ref{SI-sect:Empircal_app}.

%***DAN: I see many instances of IGR have been replaced by GWR, but there are some IGRs remaining, e.g., 
%on the figures.

@@

%***DAN: I think ATA rescue and ATA exclusion are mislabeled in the legend of Fig. 4. I think they got switched. Please check and correct if I am right.

@@

%***DAN: Jasmin, I just want to point out the below paragraph was quite well structured as it was 
%before I edited it. I only made changes on the sentence level. Nice job! Go aead and delete this 
%comment once you read it.
We produced and applied a new extension to MCT that 
quantifies the impact of ATAs on species coexistence and the maintenance of biodiversity. 
Building upon a recent, simulation-based approach to MCT \citep{Ellner_2016, Ellner_2019},
we decomposed storage effects into two new mechanisms, the contribution of ATAs to coexistence,
and the contribution of $EC$ correlation \emph{per se}.
Applications of our extended theory to the lottery model and to models based on a laboratory 
system of competing diatoms demonstrated that  
ATA contributions to storage effects and to species coexistence and the maintenance of biodiversity
can often be comparable in magnitude to the contributions of previously recognized mechanisms.
Our results highlight that common methods of correlation and covariance fail to include 
important information on the structure of relationships between variables, and our results add
coexistence and biodiversity maintenance to the list of ecological theories and phenomena on which 
ATAs have been shown to have an important impact \citep{Ghosh_2020_AER, Ghosh_2020_Eco, Ghosh_2020_ecolevol, Ghosh_2021}.  
Though storage effects are typically described as the contribution
of $EC$ \emph{covariance} to coexistence \citep{Chesson_2012, Ellner_2016, Ellner_2019},
our results suggest that storage effects should be redefined to constitute the 
contribution of \emph{associations} between $E$ and $C$, including both ATA effects and
correlation \emph{per se}; the prior terminology, \emph{covariance}, is typically construed 
as reflecting the linear relationship between variables, ignoring complexities of the association
such as ATAs.

@@

%We note that our expansion, like the decomposition of \cite{Ellner_2019} is generalizable to multi-species and multi-driver models. Because asymmetric tail associations exist in many ecological variables \citep{Ghosh_2020_AER}, the more specified approach of considering unique drivers of coexistence rather than E and C and the ability to consider more than two drivers, could produce interesting results?? 
%More species and more drivers= more things that can be related
Our further decomposition and quantification of ATA contribution to coexistence can and 
should also be applied to multi-species models. 
The inclusion of multiple species introduces a multitude of relationships between species'
environmentally dependent parameters that could possibly be asymmetrically associated. 
The greater the number of species the greater the possibilities of ATAs being more important 
to their coexistence.

%notation in ellner one could also use partial sharp 
%reread ellner paper 
%make three different biv dis that are standard normal copula and then pair with a shuffles third
%use to get rbar and subtract from rbar of XY#Z,

>>>>

%***DAN: Go through and make sure all citation callouts list papers in date order, oldest to newest,
%except when multiple pubs by the same first author are listed they are all together.
%***JASMIN: Done

@@

@@

%***DAN: I think we should replace the terminology "invasion growth rate" and "IGR" by "growth when rare" and "GWR",
% throughout the paper. Has to be done in the main text and sup mat and the figures themselves.
% Need also search for "invader" since sometimes we use that word. Also think of any other related terminology that needs to 
% be altered as part of the overall change. 
%***JASMIN: Done in text. Need to finish change in figures.

@@

%G4
To achieve goal G4 from the Introduction, i.e., to provide an empirical example of species 
coexistence which highlights the role ATAs can play, we use a laboratory chemostat
system of freshwater diatoms explored by \cite{Descamps_2005}.
Those authors used measurements of growth rates of \emph{Fragilaria crotonensis} and 
\emph{Cyclotella pseudostelligera} over a range of temperatures to parametrize a resource
competition model. The model is 
\begin{align}
\frac{dS}{dt} &= D(S_0 - S) - Q_1x_1 \frac{V_1 S}{K_1 + S} - Q_2 x_2 \frac{V_2 S}{K_2 + S} \\
\frac{dx_j}{dt} &= x_j \frac{V_j S}{K_j+S}-D x_j, 
\end{align}
for $j=1,2$, where $S$ is a extracellular silicate concentration in the chemostat, $x_1$ is the population density
of \emph{F. crotonensis} and $x_2$ is that of \emph{C. pseudostelligera}, 
$D$ is the chemostat outflow rate, $S_0$ is concentration of silicates in the chemostat
inflow, the $Q_j$ are amounts of silicate per cell, the $V_j$ are maximum population growth rates, and
the $K_j$ are half-saturation constants for the dependence of growth rates on nutrient concentration.
The temperature-dependent parameters $Q_j$, $V_j$, and $K_j$ were obtained by 
\cite{Descamps_2005} through batch experiments and curve fitting or interpolation.
\cite{Descamps_2005} showed that permanent coexistence of the two species occurs when 
temperature fluctuates periodically, 
$\theta(t)=\theta_0 +  a\sin(2\pi t/P)$, 
with parameters $\theta_0 = 18^\circ$C, $a = 6^\circ$C, $P = 60$d. 
\cite{Ellner_2016, Ellner_2019} broke down coexistence into contributing mechanisms
according to (\ref{eq:previous_decomp}) for these same values of
$\theta_0$, $a$ and $P$. We further decomponse
coexistence according to (\ref{eq:full_decomp}), and we do so for ranges of values
of $\theta_0$, $a$ and $P$. In contrast to the lottery model case, for which simplifying assumptions
made it sufficient to decompose only the GWR of species 1, for this
%***DAN: Please search for "growth when rare" and replace with GWR (except for the first instance,
%where GWR is defined). If one defines an acronym, one then has to use it.
%***JASMIN: Done
example we considered $\olr_{i \bs i} -q_{ij}\olr_{j \bs i}$ for both $i=1$, $j=2$ and for
$i=2$, $j=1$. Further details of the model setup are in SI section \ref{SI-sect:diatom_model}
and specifics of how our decomposition extends to this empirical case are in 
SI section \ref{SI-sect:Empircal_app}.

%***DAN: I see many instances of IGR have been replaced by GWR, but there are some IGRs remaining, e.g., 
%on the figures.
%***JASMIN: Figure 3 and 4 changed

@@

%***DAN: I think ATA rescue and ATA exclusion are mislabeled in the legend of Fig. 4. I think they got switched. Please check and correct if I am right.
%***JASMIN: Fixed.

@@

We produced and applied a new extension to MCT that 
quantifies the impact of ATAs on species coexistence and the maintenance of biodiversity. 
Building upon a recent, simulation-based approach to MCT \citep{Ellner_2016, Ellner_2019},
we decomposed storage effects into two new mechanisms, the contribution of ATAs to coexistence,
and the contribution of $EC$ correlation \emph{per se}.
Applications of our extended theory to the lottery model and to models based on a laboratory 
system of competing diatoms demonstrated that  
ATA contributions to storage effects and to species coexistence and the maintenance of biodiversity
can often be comparable in magnitude to the contributions of previously recognized mechanisms.
Our results highlight that common methods of correlation and covariance fail to include 
important information on the structure of relationships between variables, and our results add
coexistence and biodiversity maintenance to the list of ecological theories and phenomena on which 
ATAs have been shown to have an important impact \citep{Ghosh_2020_AER, Ghosh_2020_Eco, Ghosh_2020_ecolevol, Ghosh_2021}.  
Though storage effects are typically described as the contribution
of $EC$ \emph{covariance} to coexistence \citep{Chesson_2012, Ellner_2016, Ellner_2019},
our results suggest that storage effects should be redefined to constitute the 
contribution of \emph{associations} between $E$ and $C$, including both ATA effects and
correlation \emph{per se}; the prior terminology, \emph{covariance}, is typically construed 
as reflecting the linear relationship between variables, ignoring complexities of the association
such as ATAs.

@@

%We note that our expansion, like the decomposition of \cite{Ellner_2019} is generalizable to multi-species and multi-driver models. Because asymmetric tail associations exist in many ecological variables \citep{Ghosh_2020_AER}, the more specified approach of considering unique drivers of coexistence rather than E and C and the ability to consider more than two drivers, could produce interesting results?? 
%More species and more drivers= more things that can be related
The extended decomposition of storage effects and quantification of ATA contribution to coexistence can and 
should also be applied to multi-species models. 
The inclusion of multiple species introduces a multitude of relationships between species'
environmentally dependent parameters that could possibly be asymmetrically associated. 
The greater the number of species the greater the possibilities of ATAs being more important 
to their coexistence.

The probability of ATAs being important for the coexistence of species increases as more variables are included in the system, 
either through multi-species systems or multi-driver systems (coexistence drivers beyond just E and C). 
When multiple species included, the interaction between different species' E's and C's introduces more relationships 
to consider and could increase the relative importance of ATA contribution to the species' coexistence. 
Likewise, coexistence between species can also depend on more than two drivers \citep{Ellner_2019}. 
%notation in ellner one could also use partial sharp 
%reread ellner paper 
%make three different biv dis that are standard normal copula and then pair with a shuffles third
%use to get rbar and subtract from rbar of XY#Z,
